---
title: "8: Data Visualization Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## Opening discussion
Effective data visualization depends on purposeful choices about graph types. The ideal graph type depends on the type of data and the message the visualizer desires to communicate. The best visualizations are clear and simple. My favorite resource for data visualization is [Data to Viz](https://www.data-to-viz.com/), which includes both a decision tree for visualization types and explanation pages for each type of data, including links to R resources to create them. Take a few minutes to explore this website.

## Set Up

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)
library(ggridges)

PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.chem.nutrients.gathered <-
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
class(EPAair$Date)
class(PeterPaul.chem.nutrients$sampledate)
```

## ggplot

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_density_ridges
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot and Line Plot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot, ggplot code, draw from EPAair, basic aes must have x AND y, + tells it what kind of plot to make (+ instead of comma to go to next line!), time always goes on x axis
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

#to specific where plot shows up (in rmd file or in the "plots"), click gear/setting menu next to "Knit"

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) #shows up in environment
print(O3plot) #shows up in pdf 

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone, color = "blue"))
# do no want color to be within the actual aes layer, so put it outside parenthesis
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue")
print(O3plot2)

# Add additional variables
# want to make two different years to be different shape
# color is now coded to the site name, rather than specifying what you want the actual color of the plot to be
# shape aes has to be categorical vector, so calling year a factor in plot but still an integer in data frame
PMplot <- 
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year), color = Site.Name)) +
  geom_point()
print(PMplot)
#too many site to use color aes effectively

# Separate plot with facets
# same as above EXCEPT pulled out color aes, vars = variable that it's pulling from, nrow = number of rows you want to pop up, same with ncol
PMplot.faceted <-
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)
print(PMplot.faceted)

# Filter dataset within plot building and facet by multiple variables
# subset takes subset of (this datafram, where site name is either one of these things)
# can also say (EPAair, Site.Name %in% c("Clemmons Middle", " Leggett", "Bryson City))
# facet grid allows you to facet acorss two different variable (very sensitive to order of variables)

PMplot.faceted2 <- 
  ggplot(subset(EPAair, Site.Name == "Clemmons Middle" | Site.Name == "Leggett" |
                  Site.Name == "Bryson City"), 
         aes(x = Month, y = PM2.5)) + 
  geom_point() +
  facet_grid(Site.Name ~ Year) 
print(PMplot.faceted2)

# Plot true time series with geom_line
# time series vs true time series: true time series going continuous from day to day in a linear manner, has to be in the same spatial construct
# best for one site at a time (can use colors to plot multiple sites on same graph)
# geom_point shows actual points between line, 
# with large gaps will still connect from point to point unless adding NAs
PMplot.line <- 
  ggplot(subset(EPAair, Site.Name == "Leggett"), 
         aes(x = Date, y = PM2.5)) +
  geom_line() +
  geom_point()
print(PMplot.line)

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
# Plot surface temperatures by day of year. 
# Color your points by year, and facet by lake in two rows.
SurfaceTempPlot <- 
  ggplot(subset(PeterPaul.chem.nutrients, depth == 0),
         aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2)
print(SurfaceTempPlot)


#2. 
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5
DateTempPlot <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_point(aes(x = sampledate, y = temperature_C, color = depth),
             size = 0.5)
print(DateTempPlot)
# not a true time series because in two different lakes
```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes, xlim = x limits, ylim = y limits
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)
# trends: on the whole, DO mostly stable no matter the irradiance EXCEPT at really low depths where there is no light because there is no photosynthesis at that level

# Depth in the fields of limnology and oceanography is on a reverse scale (0 at top)
tempvsdepth <- 
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) + #draw line of best fit, lm = linear model, shading is confidence of the line (not data), remove with se = false
  geom_abline(aes(slope = 16, intercept = 0)) # slope = 16 from Redfield ratio
print(NvsP)

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by latitude 
# Make the points 50 % transparent
# Add a line of best fit for the linear regression of these variables.

  
```

### Plotting continuous vs. categorical variables
A traditional way to display summary statistics of continuous variables is a bar plot with error bars. Let's explore why this might not be the most effective way to display this type of data. Navigate to the Caveats page on Data to Viz (https://www.data-to-viz.com/caveats.html) and find the page that explores barplots and error bars. 

What might be more effective ways to display the information? Navigate to the boxplots page in the Caveats section to explore further.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill" instead of "color"?  fill shades in the middle of the box and there are a lot of conc near 0 so you can't really see it, "color" colors the dots
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient, fill = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge", dotsize = 2) #every single measurements is a specific dot (think of the cells!)
print(Nutrientplot4)

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset, have to make a separate layer for each column!
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "darkred") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "blue") +
  geom_freqpoly(aes(x = no23), color = "royalblue") +
  geom_freqpoly(aes(x = po4), color = "red") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.chem.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Frequency polygons have the risk of becoming spaghetti plots. 
# See https://www.data-to-viz.com/caveat/spaghetti.html for more info.

# Ridgeline plot
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(y = nutrient, x = concentration)) +
  geom_density_ridges(aes(fill = lakename), alpha = 0.5) #alpha is transparency piece
print(Nutrientplot6)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create several types of plots depicting PM2.5, divided by year. 
# Choose which plot displays the data best and justify your choice. 


```
