---
title: "Assignment 2: Coding Basics"
author: Hannah Smith
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1.counting by 4 to get to one hundred
question1 <- seq(1, 100, 4) 
#2. finding the mean and median of the "question 1" sequence
mean(question1)
median(question1) 
#3. determining if the mean is greater than the median (it is not)
mean(question1) > median(question1)
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
student <- c("Kitiara" , "Riverwind" , "Goldmoon" , "Tika") #character vector
score <- c(95, 83, 68, 45) #numerical vector
pass <- c(TRUE, TRUE, TRUE, FALSE) #logical vector
testscores <- data.frame(student, score, pass)
names(testscores) <- c("Student", "Score", "Passed")
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: Matrices contain the same type of element, while data frames can include different vector types (characer, numberical, logical).

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
grade_func1 <- function(x) { 
  if ( x >= 50 ) {
  print ("Pass")
    }
 else {
    print ("Fail")
 }
} #using the 'if' and 'else' function to attempt to see the passing and failing scores; does not work

grade_func2 <- function(x) {
  ifelse(x >= 50, "Pass" , "Fail") 
} #using the 'ifelse' function to determine the passing and failing scores; does work

grade_func1(score)

grade_func2(score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' worked because 'if' and 'else' can only read one object, which it cannot do because the scores are more than one object.
